Reviving the Lost Art of \\Combinator Graph Reduction

### 2023-01-30 experiments/superg/doc/paper/paper.tex
Resolve more TODOs.

We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
Better still, conversion to combinators removes all bound variables.
The runtime need not concern itself with passing arguments, this is all handled by the combinators.
The compiled program is executed using graph reduction, a technique that is also common in other reduction engines\cite{kieburtz_g-machine_1985,fairbairn_tim_1987}.

@@

\subsection{Strict algorithm}
This is based on the rules in Figure 6 of Kiselyov's paper.
We call it \textit{strict} it does not have lazy weakening: most notably it lacks the K-optimization required for full laziness.

The most naive implementation is to perform the type checking, then use the built context to apply the elimination rules and the translated program.
As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not actually care what elements are in the context, just that we keep track of the size of the context.
This lets us express compilation as a recursive function, shown in Figure~\ref{fig:strict}.

@@

There are two cases for $\lambda e$, and this is precisely where the number of elements in the context matters:
if the context is empty, the first rule applies, otherwise, we use the second one.
The semantic function may be defined as in Figure~\ref{fig:strict-sem}.

@@

These simplified functions can be implemented directly and efficiently in a conventional programming language.

\subsection{Lazy weakening}
Adding lazy weakening requires that we track which items in the context are ignored.
We change our context representation from an integer to a list of boolean values.
In the new compilation functions below, we use a shorthand $t$ for the value \textit{true} and $f$ for \textit{false}.
We always match the last element of lists and represent the remainder of the list with $\Gamma$.
Appendix~\ref{app:kiselyov} contains the new compilation function $\lazy$ (Figure~\ref{fig:lazy}).

We have introduced yet another case for $\lambda e$, this time to distinguish between $t$ or $f$ as the last element in the context of the inner expression.
There is also the new function $\sqcup$, whose definition is given in Figure~\ref{fig:lazy-ctx}.

In our implementation, we unroll the recursion into a while loop and repeatedly pop elements from the two arrays until both are empty.
This builds the new context back to front, so as a final step, the array is reversed.

Alternatively, an implementation can pad the shorter list (at the start!) with $f$ and zip them.
This is particularly convenient for languages that provide a zip function with a default value\footnote{such as \href{https://docs.python.org/3/library/itertools.html\#itertools.zip\_longest}{\texttt{ziplongest}} in Python.}.

Our semantic function now has to handle more cases, and thus is somewhat larger.
See Figure~\ref{fig:lazy-sem} for the full definition.

The first 4 cases are direct translations from the previous definitions, the others are new additions to handle ignored elements.
The last 2 cases are not included in the original paper, but they are necessary to make the pattern matching exhaustive, and are present in the reference OCaml implementation.

@@

The case $(\emptyset \models V) \coprod (\emptyset \models V) $ is impossible in this translation, implementations may throw an error or leave it as undefined behavior.
\textbf{TODO:} Why is this an impossible case?

@@

\subsection{Linear algorithm}
The implementation of the linear algorithm is much simpler because it is based on the $\strict$ algorithm.
The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$.
For the full definition, see Figure~\ref{fig:linear}.

In figure~\ref{fig:linear-sem} we give the corresponding semantic function, which is no longer recursive, making this algorithm linear.

\textbf{TODO: } Can we add lazy weakening and/or eta optimization?

@@

\begin{figure}
    \includegraphics[width=.8\columnwidth]{add_step1}
    \centering
    \caption{
        Graph of $+ \ 1 \ (K \ 2 \ 3)$ after the second argument has been added to the stack.
    }
    \label{fig:add_step1}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=.8\columnwidth]{add_step2}
    \centering
    \caption{
        Graph of $+ \ 1 \ (K \ 2 \ 3)$ after the second argument has been reduced.
    }
    \label{fig:add_step2}
\end{figure}

@@

\section{TIGRE}
\label{sec:tigre}
The TIGRE engine, first presented in~\cite{koopman_fresh_1989}, is a combinator graph reduction engine like Miranda.
Originally designed for the VAX instruction set, our implementation targets the x86-64 instruction set but mirrors the TIGRE design otherwise.
The TIGRE paper focuses on key differences from Miranda, the elimination of tags and directly executable graph nodes, so where the paper omits implementation details we take the same approach as our reference implementation based on Miranda.

@@

\begin{figure}
    \includegraphics[width=.2\columnwidth]{1_plus_2_miranda}
    \centering
    \caption{
        Graph of $1 + 2$ in the Miranda-style engine.
        Tags omitted for brevity.
    }
    \label{fig:1_plus_2_miranda}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=.46\columnwidth]{1_plus_2_tigre}
    \centering
    \caption{
        Graph of $1 + 2$ in the TIGRE-style engine.
    }
    \label{fig:1_plus_2_tigre}
\end{figure}

@@

\textbf{TODO: Maybe add a step-by-step reduction example?}

@@

It comes as no surprise that the shortest translation is also the fastest to execution, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
We see a similar pattern for the Ackermann function, for which results may be found in appendix~\ref{app:ack}.
This suggests that while the semantic translation algorithms from~\cite{kiselyov_lambda_2018} reduce the size of the combinator expression, they reduce runtime even further.

@@

\section{Conclusion}
% Repeat the main question / objective of your research
Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
We have attempted this by using a novel semantic translation algorithm to compile lambda calculus programs into combinator expressions~\cite{kiselyov_lambda_2018}, and by implementing combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.

@@

% Give the answer to the question
Our port of the TIGRE~\cite{koopman_fresh_1989} engine runs natively on stock x86-64 hardware and executes our benchmark program with fewer instructions and data accesses than our Miranda-style engine.
% Why is this the answer? Present the main arguments
Unfortunately, its performance still lags behind the reference implementation because its inherent frequent writes to executable code sections trigger expensive invalidations of the instruction cache.
We conclude that the TIGRE design is a poor fit for architectures with split instruction and data caches, though some concepts like \texttt{LIT} combinators and tag elimination may still prove effective.

>>>>

We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
Better still, conversion to combinators removes all bound variables.
The runtime need not concern itself with passing arguments, this is all handled by the combinators.

\textbf{TODO: } Combinators $B$ and $C$.

@@

\subsection{Strict algorithm}
The compilation function below, which we denote $\strict$, is based on the rules in Figure 6 of Kiselyov's paper~\cite{kiselyov_lambda_2018}.
We call it \textit{strict} it does not have lazy weakening: most notably it lacks the K-optimization required for full laziness~\cite{spj_impl}.

Kiselyov describes the algorithm as a type system with an operational semantics.
The type system describes how to assign a type, or context, to a lambda calculus term.
The operational semantics define how a typed term may be converted into combinatory logic.
A naive implementation thus requires two phases: an initial typing step, followed by a translation step.
As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not care what elements are in the context, just that we keep track of the size of the context.
This lets us express compilation as a recursive function, shown in Figure~\ref{fig:strict}.

@@

There are two cases for $\lambda e$, and this is precisely where the number of elements in the context matters:
if the context is empty, the first rule applies, otherwise, we use the second one.
Figure~\ref{fig:strict-sem} defines the semantic function.

@@

These functions are equivalent to Kiselyov's type system and can be implemented directly and efficiently in a conventional programming language.

\subsection{Lazy weakening}
The addition of Kiselyov's \textit{Lazy Weakening} requires that we track which items in the context are ignored.
The context representation changes from an integer to a list of boolean values.
In the new compilation function $\lazy$, shown in Figure~\ref{fig:lazy}, we use the shorthand $t$ for the value \textit{true} and $f$ for \textit{false}.
We always match the last element of lists and denote the remainder of the list by $\Gamma$.

We have introduced yet another case for $\lambda e$, this time to distinguish between $t$ or $f$ as the last element in the context of the inner expression.
There is also the new function $\sqcup$ to merge the left and right contexts of a function application term, whose definition is given in Figure~\ref{fig:lazy-ctx}.
Merging follows the natural rule that if either of the two contexts requires a particular item, the merged context also requires it.

In our implementation, we unroll the recursion into a while loop and repeatedly pop elements from the two arrays until both are empty.
This builds the new context back to front, so as a final step, the array is reversed.
Alternatively, an implementation can pad the shorter list (at the start!) with $f$ and zip them.
This is particularly convenient for languages that provide a zip function with a default value\footnote{such as \href{https://docs.python.org/3/library/itertools.html\#itertools.zip\_longest}{\texttt{ziplongest}} in Python.}.

Our definition of the semantic function now has to handle more cases, and thus is somewhat larger.
See Figure~\ref{fig:lazy-sem} for the full definition.
The first four cases are direct translations from the previous definitions, the others are new additions to handle ignored elements.
The last two cases are not included in the original paper, but they are necessary to make the pattern matching exhaustive, and are present in the reference OCaml implementation\footnote{From~\cite{kiselyov_lambda_2018}: \url{https://okmij.org/ftp/tagless-final/skconv.ml}}.

@@

The case $(\emptyset \models V) \coprod (\emptyset \models V) $ is impossible in this translation:
it represents the application of a value to itself, e.g. $\lambda x. x\ x$.
For such an $x$, we cannot assign a proper type.
Implementations may throw an error or leave it as undefined behavior.

@@

\subsection{Linear algorithm}
The implementation of the linear algorithm is much simpler because it is based on the $\strict$ algorithm.
The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$, shown in Figure~\ref{fig:linear-sem}.
For the full definition, see Figure~\ref{fig:linear}.

The main achievement is that the semantic function is no longer recursive.
The $\linear$ algorithm visits each term exactly once, so arguing its linearity becomes an almost trivial exercise.

We leave the addition of full laziness and $\eta$-optimization to the $\linear$ algorithm as future work.

@@

\begin{figure}
    \includegraphics[width=.6\columnwidth]{add_step1}
    \centering
    \caption{
        Graph of $+ \ 1 \ (K \ 2 \ 3)$ after the second argument has been added to the stack.
    }
    \label{fig:add_step1}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=.6\columnwidth]{add_step2}
    \centering
    \caption{
        Graph of $+ \ 1 \ (K \ 2 \ 3)$ after the second argument has been reduced.
    }
    \label{fig:add_step2}
\end{figure}

@@

\section{TIGRE}
\label{sec:tigre}
The TIGRE engine, first presented in~\cite{koopman_fresh_1989}, is a combinator graph reduction engine like Miranda.
Originally designed for the VAX instruction set, our implementation targets the x86-64 instruction set but mirrors the TIGRE design otherwise.
The TIGRE paper focuses on key differences from Miranda, the elimination of tags and directly executable graph nodes, so where the TIGRE paper omits implementation details we take the same approach as our reference implementation based on Miranda.

@@

\begin{figure}
    \includegraphics[width=.17\columnwidth]{1_plus_2_miranda}
    \centering
    \caption{
        Graph of $1 + 2$ in the Miranda-style engine.
        Tags omitted for brevity.
    }
    \label{fig:1_plus_2_miranda}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=.38\columnwidth]{1_plus_2_tigre}
    \centering
    \caption{
        Graph of $1 + 2$ in the TIGRE-style engine.
    }
    \label{fig:1_plus_2_tigre}
\end{figure}

@@

%\textbf{TODO: Maybe add a step-by-step reduction example?}

@@

It comes as no surprise that the shortest translation is also the fastest to execute, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
We see a similar pattern for the Ackermann function, for which results may be found in appendix~\ref{app:ack}.
This suggests that while the semantic translation algorithms from~\cite{kiselyov_lambda_2018} reduce the size of the combinator expression, they reduce runtime even further.

@@

\section{Conclusion}
% Repeat the main question / objective of your research
Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
We used a novel semantic translation algorithm to compile lambda calculus programs into smaller combinator expressions~\cite{kiselyov_lambda_2018} and have implemented modern combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.

@@

% Give the answer to the question
Our port of the TIGRE~\cite{koopman_fresh_1989} engine runs natively on stock x86-64 hardware and executes our benchmark program with fewer instructions and data accesses than our Miranda-style engine.
% Why is this the answer? Present the main arguments
Unfortunately, its performance still lags behind the Miranda-style engine because its inherent frequent writes to executable code sections trigger expensive invalidations of the instruction cache.
We conclude that the TIGRE design is a poor fit for architectures with split instruction and data caches, though some concepts like \texttt{LIT} combinators and tag elimination may still prove effective.
